Effective collector efficiency of sparse and dense arrays of emergent cylindrical collectors in laminar-turbulent transitional flows

### 2020-12-16 aa workingdraft/draft.tex
Updates from Overleaf

% Flume diagram figure
\begin{figure}[h]
\includegraphics[width=5in]{../pics/flume_with_sedtraps.png}
\centering
\caption{The Ecogeomorphology flume. Not all measurements are to scale. \textbf{(a)} Photograph of the test section (center-left) and pump (right). \textbf{(b)} Conceptual diagram of the flume as seen from above. Labeled parts are: 1) pump, 2) magnetic flowmeter, 3) test section, and 4) honeycomb flow collimator. Arrows indicate direction of flow. Green points represent the inlets for the peristaltic pumps sampling suspended particle concentration. Red areas represent the settling traps. \textbf{(c)} A side view of the open-channel part of the system.}
\label{fig:floorplan}
\end{figure}

@@

Within the open channel, we designated a "test section" 1.95 m in length where collectors would be installed. This test section and the surrounding area were instrumented with the following devices:
\begin{enumerate}
     \item A flat-bedded array positioned flush with the neighboring channel bed, containing vertical, emergent collector stems, which were spaced uniformly in an unpatterned manner
    \item Two battery-operated peristaltic pumps (Cole-Parmer, Vernon Hills, IL) used to sample suspended particle concentration via three hose inlets each (inside diameter = 3.1 mm), which were suspended at a range of heights from the channel bed (5; 14; 27 cm); these were positioned 50 cm upstream and downstream of the test section 
   \item Settling traps (n = 9) with 2.5 cm circular openings flush with the bed and collection filters (Whatman GF/F) on perforated filter holders recessed in a 5 cm deep cylindrical cavity (trap Reynolds number and aspect ratio chosen to minimize bias \cite{butman1986sediment}); these were interspersed among the collectors in a grid-like pattern
\end{enumerate}

We used 1/8" (0.3175 cm) cylindrical wooden dowels as collectors, which were covered in silicone grease (Chemplex 710, Fuchs Petrolub, Mannheim, Germany) in order to retain impacted particles. Crushed walnut shell was used as a substitute for suspended sediment because it has surficial and physical properties comparable to the organic-rich types common in wetlands \cite{muller2017experiments, jenzer2015sediment, redding2006particle}. We used WF5-200 grade (Composition Materials Co., Milford, CT), which passes entirely through a \#60 sieve (\SI{250}{\micro\metre}), and which we determined has a average particle diameter ($d_{50}$) of \SI{25.2}{\micro\metre} based on measurements using a laser-scattering-based instrument (LISST-Portable|XR, Sequoia Scientific, Bellevue, WA; See Appendix A). We also empirically estimated the particle density for this specific walnut shell flour to be 1.53 \SI{}{\gram/\centi\metre\cubed}.

@@

Before each experiment, the flume was filled with tap water to 0.4 m depth in the test section, which was enough to fully submerge the pump and almost the entirety of the duct length. At this depth, the volume of water in the entire flume was approximately 2.43 m$^3$ (See Appendix B).  A Nortek Vectrino Profiler ADV (acoustic Doppler velocimeter; Nortek AS, Vangkroken 2, 1351 Rud, Norway) was used to calibrate the pump to generate channel mean flow velocities (2.0; 4.0; 6.0 \SI{}{\centi\metre/\second}) within the pump's capabilities and corresponding to the range of interest for $\Rey_c$. At these flow rates (0.0048; 0.0096; 0.0144 \SI{}{\metre\cubed/\second}), circulation times for the entire system were 506 s, 253 s, and 169 s, respectively. The drag caused by collectors was not found to detectably affect upstream or downstream flow velocities.

Settling traps were installed immediately before each run, then a slurry consisting of 200 g of crushed walnut shell suspended in 15.1 L of tap water was added using a spigot calibrated to finish draining after a period roughly equal to circulation time (3 minutes). Channel velocity was kept at 6 cm/s for this period across $\Rey_c$ treatments, then adjusted to the appropriate value for the treatment. This procedure was intended to make particle concentration longitudinally even throughout the system. Estimated depth-averaged starting concentration was 82.3 mg/L (\SI{53.8}{\micro\liter/\liter}).

In preliminary experiments, we determined 100 minutes was an adequate duration for experimental runs to capture the effects of settling and collection on suspended concentrations across our parameter space (See Appendix C). We sampled suspended concentration using the peristaltic pumps every 5 minutes, the first sample occurring 5 minutes after we began adding sediment to the flume. Samples measured approximately 140 ml each, and our sampling frequency resulted in 19 sampled timesteps per experiment, with six samples per timestep. Settling traps were covered with a plunger and removed at 100 minutes, so the last peristaltic pump sample occurred at 95 minutes due to personnel limitations.

Peristaltic pump samples were processed using vacuum filtration. Before filtration, quantitative glass microfiber filter papers (Whatman GF/F) were oven-dried for 24 hours at \SI{40}{\celsius} to remove any moisture, weighed, and assigned to a flume sample. The volume of flume water passed through the filter was measured, then the drying and weighing process was repeated to estimate the mass of sediment on the filter. Concentrations were recorded in \SI{}{\milli\gram/\liter}. Settling traps, which also used filter papers, were dried and weighed before and after the run in a similar fashion, to estimate the mass of settled particles they contained.

@@

To calculate $k_s$, we began by calculating the average mass of sediment that settled per settling trap over the duration of the experiment (100 minutes). We then scaled this measurement up from the area of a settling trap's horizontal opening (5.07 $\times 10^{-4}$ m$^2$) to the total test section bed area (1.17 m$^2$). The resulting estimate of the sediment mass settled in the test section ($m_s$) was divided by the total mass removed from suspension during a run to calculate the proportion of the decrease due to settling. This proportion was multiplied by the total exponential decay coefficient, $k$, to calculate $k_s$:

@@

\noindent which was evaluated for each experimental configuration separately. Bold font signifies vector variables, and $\boldsymbol{U}$ is the random error term of the model. We used residual plots to verify that this transformed model exhibited uniformity of residuals (See Appendix D). The standard error of this $k$ term and the standard deviation of $m_s$ estimates from settling traps were propagated through Eqs. (\ref{eq:ks}) and (\ref{eq:eta}), for both control and treatment runs, and were ultimately used to calculate uncertainty estimates for ECE.

@@

We used a Monte Carlo approach to assess the significance of the effects on ECE of $\Rey_c$ and collector density, the latter parameterized as solid volume fraction ($\phi_c$) in the interest of comparability with other studies. For each combination of $\Rey_c$ and $\phi_c$, 30,000 random values were drawn from a normal distribution whose mean equals that combination's empirically estimated ECE and whose standard deviation equals its standard error (SE). These values were log-transformed, since we expected the relationships to fit a power-law model. For each stratum of a given explicative variable, sets (triplets) with one randomly sampled value from each of the strata of the other explicative variable were regressed using ordinary least squares, producing a roughly normal distribution of coefficients (n = 30,000) for each level of both variables (6 total). The distributions of slopes were used in two-tailed t-tests (null hypothesis: $\mu = 0$; $\alpha = 0.05$) to determine the significance of relationships.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendix
\section{Secondary methodology}
\unskip
\subsection{Particle size}
\begin{figure}[H]
\centering
\includegraphics[width=5in]{../pics/wf5-200sizedist.png}
\caption{Walnut shell flour (WF5-200) particle size distribution, as measured by LISST-XR|Portable. (\textbf{a}) Probability density function, with mode (\SI{32.55}{\micro\metre}) indicated (\textbf{b}) Cumulative density function, with 50th (\SI{25.20}{\micro\metre}) and 84th (\SI{57.53}{\micro\metre}) percentiles highlighted}
\end{figure}

\subsection{Flume volume}

In order to estimate $\eta$, we need to know the collection rate for water that is being acted upon by the collectors (i.e., in the test section), which requires scaling up by the ratio between the test section volume (1.95 m $\times$ 0.6 m $\times$ 0.4 m = \SI{0.468}{m^3}) and the water volume of the flume as a whole. To calculate the latter, we drained the flume while measuring flow rate at the drain outlet over time, performed polynomial regression, and calculated a definite integral from when draining began to when the part of the flume involved in the experimental flows was empty ($V = \int_{t_{full}}^{t_{empty}}{-\frac{dV}{dt}dt}$).

\begin{figure}[H]
\centering
\includegraphics[width=5in]{../pics/wf5-200sizedist.png}
\caption{Walnut shell flour (WF5-200) particle size distribution, as measured by LISST-XR|Portable. (\textbf{a}) Probability density function, with mode (\SI{32.55}{\micro\metre}) indicated (\textbf{b}) Cumulative density function, with 50th (\SI{25.20}{\micro\metre}) and 84th (\SI{57.53}{\micro\metre}) percentiles highlighted}
\end{figure}

\subsection{Experimental run duration}

\unskip
\subsection{Heteroskedasticity remediation}

>>>>

% Flume diagram figure
\begin{figure}[h]
\includegraphics[width=5in]{../pics/flume_with_sedtraps.png}
\centering
\caption{The Ecogeomorphology flume. Not all measurements are to scale. \textbf{(a)} Photograph of the test section (center-left) and pump (right). \textbf{(b)} Conceptual diagram of the flume as seen from above. Labeled parts are: 1) pump, 2) magnetic flowmeter, 3) test section, and 4) honeycomb flow collimator. Arrows indicate direction of flow. Green points represent the inlets for the peristaltic pumps sampling suspended particle concentration. Red areas represent the sediment traps. \textbf{(c)} A side view of the open-channel part of the system.}
\label{fig:floorplan}
\end{figure}

@@

Within the open channel, we designated a "test section" 1.95 m in length where collectors would be installed. This test section and the surrounding area were instrumented with the following devices:
\begin{enumerate}
     \item A flat-bedded array positioned flush with the neighboring channel bed, containing vertical, emergent collector stems, which were spaced uniformly in an unpatterned manner
    \item Two battery-operated peristaltic pumps (Cole-Parmer, Vernon Hills, IL) used to sample suspended particle concentration via three hose inlets each (inside diameter = 3.1 mm), which were suspended at a range of heights from the channel bed (5; 14; 27 cm); these were positioned 50 cm upstream and downstream of the test section 
   \item sediment traps (n = 9) with 2.5 cm circular openings flush with the bed and collection filters (Whatman GF/F) on perforated filter holders recessed in a 5 cm deep cylindrical cavity (trap Reynolds number and aspect ratio chosen to minimize bias \cite{butman1986sediment}); these were interspersed among the collectors in a grid-like pattern
\end{enumerate}

We used 1/8" (0.3175 cm) cylindrical wooden dowels as collectors, which were covered in silicone grease (Chemplex 710, Fuchs Petrolub, Mannheim, Germany) in order to retain impacted particles. Crushed walnut shell was used as a substitute for suspended sediment because it has surficial and physical properties comparable to the organic-rich types common in wetlands \cite{muller2017experiments, jenzer2015sediment, redding2006particle}. We used WF5-200 grade (Composition Materials Co., Milford, CT), which passes entirely through a \#60 sieve (\SI{250}{\micro\metre}), and which we determined has a average particle diameter ($d_{50}$) of \SI{25.2}{\micro\metre} based on measurements using a laser-scattering-based instrument (LISST-Portable|XR, Sequoia Scientific, Bellevue, WA; See Appendix A1). We also empirically estimated the particle density for this specific walnut shell flour to be 1.53 \SI{}{\gram/\centi\metre\cubed}.

@@

Before each experiment, the flume was filled with tap water to 0.4 m depth in the test section, which was enough to fully submerge the pump and almost the entirety of the duct length. At this depth, the volume of water in the entire flume was approximately 2.43 m$^3$ (See Appendix A2).  A Nortek Vectrino Profiler ADV (acoustic Doppler velocimeter; Nortek AS, Vangkroken 2, 1351 Rud, Norway) was used to calibrate the pump to generate channel mean flow velocities (2.0; 4.0; 6.0 \SI{}{\centi\metre/\second}) within the pump's capabilities and corresponding to the range of interest for $\Rey_c$. At these flow rates (0.0048; 0.0096; 0.0144 \SI{}{\metre\cubed/\second}), circulation times for the entire system were 506 s, 253 s, and 169 s, respectively. The drag caused by collectors was not found to detectably affect upstream or downstream flow velocities.

sediment traps were installed immediately before each run, then a slurry consisting of 200 g of crushed walnut shell suspended in 15.1 L of tap water was added using a spigot calibrated to finish draining after a period roughly equal to circulation time (3 minutes). Channel velocity was kept at 6 cm/s for this period across $\Rey_c$ treatments, then adjusted to the appropriate value for the treatment. This procedure was intended to make particle concentration longitudinally even throughout the system. Estimated depth-averaged starting concentration was 82.3 mg/L (\SI{53.8}{\micro\liter/\liter}).

We determined 100 minutes was the necessary duration for experimental runs to capture the effects of settling and collection on suspended concentrations across our parameter space (See Appendix A4). We sampled suspended concentration using the peristaltic pumps every 5 minutes, the first sample occurring 5 minutes after we began adding sediment to the flume. Samples measured approximately 140 ml each, and our sampling frequency resulted in 19 sampled timesteps per experiment, with six samples per timestep. sediment traps were covered with a plunger and removed at 100 minutes, so the last peristaltic pump sample occurred at 95 minutes due to personnel limitations.

Peristaltic pump samples were processed using vacuum filtration. Before filtration, quantitative glass microfiber filter papers (Whatman GF/F) were oven-dried for 24 hours at \SI{40}{\celsius} to remove any moisture, weighed, and assigned to a flume sample. The volume of flume water passed through the filter was measured, then the drying and weighing process was repeated to estimate the mass of sediment on the filter. Concentrations were recorded in \SI{}{\milli\gram/\liter}. sediment traps, which also used filter papers, were dried and weighed before and after the run in a similar fashion, to estimate the mass of settled particles they contained.

@@

To calculate $k_s$, we began by calculating the average mass of sediment that settled per sediment trap over the duration of the experiment (100 minutes). We then scaled this measurement up from the area of a sediment trap's horizontal opening (5.07 $\times 10^{-4}$ m$^2$) to the total test section bed area (1.17 m$^2$). The resulting estimate of the sediment mass settled in the test section ($m_s$) was divided by the total mass removed from suspension during a run to calculate the proportion of the decrease due to settling. This proportion was multiplied by the total exponential decay coefficient, $k$, to calculate $k_s$:

@@

\noindent which was evaluated for each experimental configuration separately. Bold font signifies vector variables, and $\boldsymbol{U}$ is the random error term of the model. We used residual plots to verify that this transformed model exhibited uniformity of residuals (See Appendix A3). The standard error of this $k$ term and the standard deviation of $m_s$ estimates from sediment traps were propagated through Eqs. (\ref{eq:ks}) and (\ref{eq:eta}), for both control and treatment runs, and were ultimately used to calculate uncertainty estimates for ECE.

@@

Lastly, we used a Monte Carlo approach to assess the significance of the effects on ECE of $\Rey_c$ and collector density, the latter parameterized as solid volume fraction ($\phi_c$) in the interest of comparability with other studies. For each combination of $\Rey_c$ and $\phi_c$, 30,000 random values were drawn from a normal distribution whose mean equals that combination's empirically estimated ECE and whose standard deviation equals its standard error (SE). These values were log-transformed, since we expected the relationships to fit a power-law model. For each stratum of a given explicative variable, sets (triplets) with one randomly sampled value from each of the strata of the other explicative variable were regressed using ordinary least squares, producing a roughly normal distribution of coefficients (n = 30,000) for each level of both variables (6 total). The distributions of slopes were used in two-tailed t-tests (null hypothesis: $\mu = 0$; $\alpha = 0.05$) to determine the significance of relationships.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendix
\section{Secondary methodology}
\unskip
\subsection{Particle Size}
\begin{figure}[H]
\centering
\includegraphics[width=5in]{../pics/wf5-200sizedist.png}
\caption{Walnut shell flour (WF5-200) particle size distribution, as measured by LISST-XR|Portable. (\textbf{a}) Probability density function, with mode (\SI{32.55}{\micro\metre}) indicated (\textbf{b}) Cumulative density function, with 50th (\SI{25.20}{\micro\metre}) and 84th (\SI{57.53}{\micro\metre}) percentiles highlighted}
\end{figure}

\subsection{Flume Volume}

In order to estimate ECE, we need to know the collection rate for water that is being acted upon by the collectors (i.e., in the test section), which requires scaling up by the ratio between the test section volume (1.95 m $\times$ 0.6 m $\times$ 0.4 m = \SI{0.468}{m^3}) and the water volume of the flume's entire wetted floorplan under experimental conditions (depth = 0.4 m in test section). To calculate the latter, we drained the flume after an experiment while measuring flow rate at the drain outlet over time, performed polynomial regression (Figure \ref{fig:flumevol}), and calculated a definite integral (Table \ref{tbl:flumevol}) from when draining began to when the part of the flume involved in the experimental flows was empty ($V = \int_{t_{full}}^{t_{empty}}{-\frac{dV}{dt}dt}$). Because polynomials of second order and above fit well, and all yielded very similar volume estimates, we felt confident using \SI{2.43}{\metre\cubed} as the flume volume in our calculations.

\begin{figure}[H]
\centering
\includegraphics[width=5in]{../pics/flumevol.png}
\caption{Polynomial models predicting the rate at which the flume drained, to be integrated to estimate the volume in circulation during experiments.}
\label{fig:flumevol}
\end{figure}

\begin{table}[H]
\caption{Polynomial model summary}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{lrr}
\toprule
\textbf{Model}&\textbf{Estimated Volume (\SI{}{\metre\cubed})}&\textbf{R$^2$}\\
\midrule
Linear       &  2.4309     &   0.80\\
Quadratic    &  2.4277     &   0.90\\
Cubic        &  2.4373     &   0.94\\
4th-order    &  2.4294     &   0.96\\
\bottomrule
\label{tbl:flumevol}
\end{tabular}
\end{table}

\subsection{Heteroskedasticity Remediation}

\subsection{Experimental Run Duration}

The U-shaped pattern in the residual plot could be attributable to the bed, walls, and other surfaces in the flume, including the collector surfaces, beginning the experiment free of particles. This means that at the beginning of the run, re-suspension from these surfaces is equal to zero. As particles adhere during the run, the absolute rate of re-suspension can be assumed to increase even if $p_r$ remains constant, resulting in a declining overall magnitude of exponential decay of suspended particle concentration. It should be noted that we observed re-entrainment from the test section bed in particular to be negligible, but this does not necessarily apply to other aspects of the flume.

To minimize the effect of changing re-suspension rates on our analyses, we sought to minimize the positive curvature in our residual plot without making substantial sacrifices in terms of model fit. We went about doing so by conducting models for an $n \times m$ grid with $n$ timesteps dropped from the beginning and $m$ from the end of all experimental runs in the dataset. We calculated a crude estimate of curvature using the second-order coefficient of a quadratic model of the residuals, and the mean and minimum R$^2$ of the set of 12 linear models used in analyses, for each point in the grid (Figure).
